Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
4

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{is designed to accept the lazy} semantics of
linearity. \ROUNDTWO{Our system achieves this via a novel combination
  of features: explicit tracking of captured linear resources through
        \emph{usage environments} (Section~\ref{sec:usage-environments});
  a distinct treatment of case
  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf});
  and a notion of irrelevant and tagged resources
  (Section~\ref{sec:irrev}), which allow for fine-grained control over
  variable usage.} 
%
To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

Linear Core makes precise the various insights discussed in the
previous section in a linear type system that we prove \ROUNDTWO{to be} sound and guarantee
linear resource usage at runtime.
This result is obtained (see Section~\ref{sec:main:metatheory}) via a soundness argument on a
call-by-need big-step operational semantics that becomes stuck when linear
variables in the runtime heap are used more than once, \ROUNDTWO{a
  technical approach that is natural in the literature \cite{}}. We show this
semantics to be bisimilar to a
Launchbury-style~\cite{10.1145/158511.158618} natural semantics. 
%\todo{O que digo aqui sobre a semantica launchbury style?}
Crucially,
Linear Core typing accepts all the semantically linear examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Additionally, we prove that multiple optimising Core-to-Core transformations
preserve linearity in Linear Core, while violating it under Core's
current type system.
%
%To the best of our knowledge, we are the first to prove optimisations preserve
%types in a non-strict linear language.
% 
Despite the focus on GHC Core, the fundamental ideas for \ROUNDTWO{typing} linearity in
our call-by-need calculus may be applied to other languages \ROUNDTWO{which
combine non-strict features or semantics with linearity}.

\ROUNDTWO{The key ideas that make Linear Core work are:}

\ROUNDTWO{
\begin{enumerate}
    \item Usage environments (Section~\ref{sec:usage-environments})
    \item Distinct treatment of case scrutinees in WHNF (Section~\ref{sec:onwhnf})
    \item Irrelevant and tagged resources (Sections~\ref{sec:irrev}~and~\ref{sec:tagging})
\end{enumerate}

\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
definition, but rather when the bound variables themselves are consumed.
Consuming a variable with some usage environment $\Delta$ equates to consuming
the variables contained in $\Delta$.
%
(2) Case expressions in WHNF may capture ambient linear resources, but these
resources can be safely used in branches since the case will not further
evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
must be treated more conservatively.
%
(3) Irrelevant resources make existing bindings unusable while forcing the use
of others. Namely, linear resources occurring in non-WHNF case scrutinees can
no longer be used in the alternatives, while the pattern variables (or the case
binder) must necessarily be used. Tagged resources follow the observation that
some variables must be used jointly or not at all.
}
%
% Linear Core's first key idea is to delay consuming
% resources until a computation that depends on those resources is effectively
% evaluated or returned. Delayed consumption is tracked by annotating relevant
% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
% The second key idea is to have two distinct rules for case
% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
% longer available but have yet to be fully consumed
% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
% resources amongst pattern-bound variables and %^as usage environments
% encode that all linear components of a pattern must be used in tandem.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
the scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

\subsubsection{Splitting and tagging fragments}\label{sec:tagging}

@@

\renewcommand{\DeltaLinearRelationLemma}{
\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{restatable}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
\end{restatable}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{restatable}
}

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context and
\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$}.

@@

\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
environment can be made into a linear variable as long as its
(irrelevant) environment is dropped from the linear context.

@@

\paragraph{Irrelevance}
%
Proof irrelevant resources can only be consumed
indirectly, essentially encoding that the (non-WHNF) case scrutinee
resources must be consumed through the case binder or the linear
pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually
reduce to WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, the two case rules must be in harmony in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.

@@

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources with respect to
typing the same expression with arbitrary resources. However, typing
a case alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly.

@@

%\subsection{Type safety\label{sec:type-safety-meta}}
% The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
% get \emph{stuck} and their evaluation preserves linearity.
We prove type safety of the Linear Core system via the standard type
preservation and progress results. As is customary, we make use of multiple
substitution lemmas, one for each kind of variable: unrestricted variables
$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
development is reported in Appendices~\ref{app:proofs}
and~\ref{app:opsem}, where type safety is established directly for the
instrumented semantics and ported to the natural semantics via a
bisimulation argument.

@@

  {\small
  \[
    \begin{array}{ll}
   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
                                  \mbox{Case of known constr.}\\
      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
            \Longrightarrow 
            %{{{
      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
      & \mbox{Case-of-case}\\
      (\lambda \y[\pi].~\llet{\xD = e}{e'})
      \Longrightarrow
      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
      %{{{
      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
                               &
                                 \mbox{Commuting let-case}\\
      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
      %{{{
     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
                                                                        & \mbox{Commuting case-let}\\
      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
                                                                      & \mbox{Commuting let-let}\\
      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
                                                                      & \mbox{Binder swap}\\
      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
      %                                                                 & \mbox{Reverse binder swap}
    \end{array}
  \]}
  \caption{Optimising transformations validated by Linear
    Core\label{fig:opttrans} (for commuting case-let we
      require $e_1$ not to capture the case binder, variables bound by the
      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
    details).} 
\end{figure}

>>>>

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{is designed to accept the lazy} semantics of
linearity. \ROUNDTWO{Our system achieves this via a novel combination
  of features: explicit tracking linear dependencies of variables
  through \emph{usage environments}
  (Section~\ref{sec:usage-environments}), codifying that lazy variable
  bindings do not consume resources upon definition, but rather when
  the variables are consumed;
  a distinct treatment of case
  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
  identifying that linear resources captured by case expressions in
  WHNF can be safely used in branches since the case will not further
  evaluate its scrutinee; and a notion of irrelevant and tagged resources
  (Section~\ref{sec:irrev}), which allow for fine-grained control over
  variable usage. Namely, linear resources occurring in non-WHNF case
  scrutinees can no longer be used in the alternative, while the
  pattern variables (or the case binder) must necessarily be used. } 
%
To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

Linear Core makes precise the various insights discussed in the
previous section in a linear type system that we prove \ROUNDTWO{to be} sound and guarantee
linear resource usage at runtime.
This result is obtained (see Section~\ref{sec:main:metatheory}) via a soundness argument on a
call-by-need big-step operational semantics that becomes stuck when linear
variables in the runtime heap are used more than once, \ROUNDTWO{a
  technical approach that is natural in the literature \cite{}}. We show this
semantics to be bisimilar to a
Launchbury-style~\cite{10.1145/158511.158618} natural semantics. 
%\todo{O que digo aqui sobre a semantica launchbury style?}
Crucially,
Linear Core typing accepts all the semantically linear examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Additionally, we prove that multiple optimising Core-to-Core transformations
preserve linearity in Linear Core, while violating it under Core's
current type system.
%
%To the best of our knowledge, we are the first to prove optimisations preserve
%types in a non-strict linear language.
% 
Despite the focus on GHC Core, the fundamental ideas for
understanding linearity in our call-by-need calculus can be readily applied to
other non-strict languages.

Linear Core's first key idea is to delay consuming
resources until a computation that depends on those resources is effectively
evaluated or returned. Delayed consumption is tracked by annotating relevant
binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for case
expressions, branching on whether the scrutinee is in Weak Head Normal
Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
longer available but have yet to be fully consumed
(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
resources amongst pattern-bound variables and %^as usage environments
encode that all linear components of a pattern must be used in tandem.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

\subsubsection{Splitting and tagging fragments}

@@

\renewcommand{\DeltaLinearRelationLemma}{
\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \s$ 
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\s$.
\end{restatable}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
If $\G; \D',\xl \vdash e :\s$
then $\G[\D/x],\xD; \D,\D' \vdash e : \s$ ($\Delta$ fresh).
\end{restatable}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \s$ iff $\G,\x[\ROUNDTWO{\emptyset}]; \D \vdash e : \s$
\end{restatable}
}

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context and
\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.}

@@

\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
environment can be made into a linear variable as long as its
(irrelevant) environment is dropped from the linear context
\ROUNDTWO{and all the occurrences of the set of variables $[\D]$ in
  usage environments in $\Gamma$ are collapsed to $x$, written $\Gamma[x/[\Delta]]$}.

@@

\paragraph{Irrelevance}
%
Irrelevant resources \ROUNDTWO{(Section~\ref{sec:irrev})} can only be consumed
indirectly, essentially encoding that the non-WHNF case scrutinee
resources must be consumed through the case binder or the linear
pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually
reduce to WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, the two case rules must be in harmony in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.

@@

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with irrelevant resources with respect to
typing the same expression with arbitrary resources. However, typing
a case alternative with irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly.

@@

%\subsection{Type safety\label{sec:type-safety-meta}}
% The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
% get \emph{stuck} and their evaluation preserves linearity.
We prove type safety of the Linear Core system via the standard type
preservation and progress results. As is customary, we make use of multiple
substitution lemmas, one for each kind of variable: unrestricted variables
$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
development is reported in \cite{techreport}, where type safety is established directly for the
instrumented semantics and ported to the natural semantics via a
bisimulation argument.

@@

  {\small
  \[
    \begin{array}{ll}
   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\[1ex]
       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\[1ex]
      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
                                  \mbox{Case of known constr.}\\[1.5ex]
      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
            \Longrightarrow 
            %{{{
      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
      & \mbox{Case-of-case}\\[1.5ex]
      (\lambda \y[\pi].~\llet{\xD = e}{e'})
      \Longrightarrow
      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\[1ex]
      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\[1ex]
      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
      %{{{
      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
                               &
                                 \mbox{Commuting let-case}\\[1ex]
      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\[1.5ex]
      %{{{
     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
                                                                        & \mbox{Commuting case-let}\\[1.5ex]
      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
                                                                      & \mbox{Commuting let-let}\\[1ex]
      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\[1ex]
      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\[1ex]
      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
                                                                      & \mbox{Binder swap}\\[1ex]
      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
      %                                                                 & \mbox{Reverse binder swap}
    \end{array}
  \]}
  \caption{Optimising transformations validated by Linear
    Core\label{fig:opttrans} (for commuting case-let we
      require $e_1$ not to capture the case binder, variables bound by the
      pattern $\rho$ and the context $E[{-}]$ -- see \cite{techreport} for
    details).} 
\end{figure}
